Generalized Bradley-Terry models

### 2025-03-20 inst/jss_hyper3_real.Rnw
more JSS house style

\begin{table}[t!]
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
\end{table}
%
We are now in a position to consider the entirety of
Table~\ref{oldfirm}.  Bespoke package function
\code{home\_draw\_away3()} converts data in the form of
Table~\ref{oldfirm} to a \code{hyper3} likelihood function following
Table~\ref{windrawloss}; it takes numerical values for the home
advantage $\lambda$ and proclivity to draw $D$:
%
<<eg_RCLF3_lf, cache=TRUE>>=
home_draw_away3(RCLF3_table, lambda=1.7, D=0.3)
@
%
Observe that the resulting likelihood function has two distinct types
of parameters: the plain Bradley-Terry strengths $\mathbf{p}=(p_R,
p_C,p_L,p_F)$, with unit sum; but also the auxiliary parameters
$\lambda,D$.  To make inferences about these parameters we calculate a
profile likelihood function: conditional on $\lambda,D$ we maximize
the likelihood over possible values of $\mathbf{p}$.
%
<<defineF>>=
f <- function(v){
    H <- home_draw_away3(RCLF3_table, lambda = v[1], D=v[2])
    maxp(H,give=1)$`Log-likelihood`
}
@
%
Thus \code{f()} gives us a likelihood function for
$\mathbf{v}=(\lambda,D)$.  We may find the evaluate using
\code{optim()}:
%
<<findmaxlambdaD, cache=TRUE>>=
maxv <- optim(par=c(2, 0.3), fn=f, control=list(fnscale = -1))$par
maxv
@
%
Then
%
<<showevaluate, cache=TRUE>>=
maxp(home_draw_away3(RCLF3_table, lambda = maxv[1], D=maxv[2]))
@
%
Such estimates have rich predictive power.  For example, if Rangers
were to play Celtic with both teams away [thus nullifying the home
  ground advantage: $\lambda=1$ in Table~\ref{windrawloss}], the
probability of win/draw/loss for Rangers would be
$\left(\frac{p_R}{(1+D)(p_R+p_C)};\frac{D}{1+D};\frac{p_C}{(1+D)(p_R+p_C)}\right)$,
estimated at $(0.39;0.27;0.34)$.
%
<<makecontourdata, echo=FALSE, cache=TRUE>>=
n <- howmany
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
@

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce~\citep{plackett1975} likelihood function for
this observation:
%
\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}
%
Note the use of multiplication between the terms, indicating
conditional independence.  Now we use the multiplicative
generalization of Bradley-Terry strengths to introduce an element of
non-independence between the terms.  There are many ways to do this,
but one simple way would be to define equivalence classes of the
competitors with each equivalence class comprising mutually supporting
runners.  The metaphor would be that a runner who has finished the
race is able to support other members of his equivalence class by
cheering his still-running teammates, boosting their performance.  The
\pkg{hyper2} package includes function \code{cheering()} which
implements this functionality.

@@

Specifically, it seems reasonable to assume that if we ask our
respondent to rank his options, it is highly probable that he will put
RB and BB consecutively (because they are essentially
indistinguishable).  Can we quantify the strength of this effect?  To
do this, we define a bespoke function \code{RB\_BB\_LF()} which
returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options.
%
<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
    )
}
@
%
Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Vector \code{ec} places the options into equivalence classes;
equivalence class 3 comprises the two buses and the other classes are
unsupported as they are of size one.  Vector \code{h} encodes the
strength of the nonindependence.  Observe that in almost every case he
ranks RB and BB consecutively.  Function \code{RB_BB_LF()} takes
argument \code{lambda} that quantifies the perceived similarity
between RB and BB in the same way as function \code{oldfirm()} above.
%
<<definefunctiono>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
@ 
%
Above, we define function \code{o()} which returns the support for a
given value of $\lambda$ at the evaluate [over permissible Bradley
  Terry strengths for the options].  We can treat \code{o()} as a
support function for $\lambda$, and then maximize it:
%
<<label=maxlikesim, cache=TRUE>>=
(osup <- optimize(o, c(10, 40), maximum=TRUE))
@ 
%
So a likelihood ratio test of the null that $\lambda=1$ would be:
%
<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- osup$objective - o(lambda = 1))
@
%
We can thus improve the hypothesis that $\lambda=1$, corresponding to
an unaltered Plackett-Luce system, by over 5.6 units of support by
dint of choosing instead $\lambda = 10.49$.  This easily satisfies
\citeauthor{edwards1992}'s two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
%
%It is straightforward to find a profile
%support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
%We see that support of $\log\lambda$ is closely quadratic over a broad
%range.
%
%\begin{figure}[htbp]
%\begin{center}
<<plotproflike, echo=FALSE>>=
if(FALSE){
    lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
    L <- sapply(lambda, o)
    plot(log(lambda), L-max(L), type="b")
    abline(h=c(0, -2))
    abline(v=0)
}
@ 
%\caption{A\label{profilesupportlambdaRBBB} profile support function
%  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
%  Vertical line indicates null of $\lambda=1$, horizontal lines show
%  maximum support and $-2$}
%\end{center}
%\end{figure}

>>>>

\begin{table}[t!]
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
\end{table}
%
We are now in a position to consider the entirety of
Table~\ref{oldfirm}.  Bespoke package function
\fct{home\_draw\_away3} converts data in the form of
Table~\ref{oldfirm} to a \code{hyper3} likelihood function following
Table~\ref{windrawloss}; it takes numerical values for the home
advantage $\lambda$ and proclivity to draw $D$:
%
<<eg_RCLF3_lf, cache=TRUE>>=
home_draw_away3(RCLF3_table, lambda=1.7, D=0.3)
@
%
Observe that the resulting likelihood function has two distinct types
of parameters: the plain Bradley-Terry strengths $\mathbf{p}=(p_R,
p_C,p_L,p_F)$, with unit sum; but also the auxiliary parameters
$\lambda,D$.  To make inferences about these parameters we calculate a
profile likelihood function: conditional on $\lambda,D$ we maximize
the likelihood over possible values of $\mathbf{p}$.
%
<<defineF>>=
f <- function(v){
    H <- home_draw_away3(RCLF3_table, lambda = v[1], D=v[2])
    maxp(H,give=1)$`Log-likelihood`
}
@
%
Thus \fct{f} gives us a likelihood function for
$\mathbf{v}=(\lambda,D)$.  We may find the evaluate using
\fct{optim}:
%
<<findmaxlambdaD, cache=TRUE>>=
maxv <- optim(par=c(2, 0.3), fn=f, control=list(fnscale = -1))$par
maxv
@
%
Then
%
<<showevaluate, cache=TRUE>>=
maxp(home_draw_away3(RCLF3_table, lambda = maxv[1], D=maxv[2]))
@
%
Such estimates have rich predictive power.  For example, if Rangers
were to play Celtic with both teams away [thus nullifying the home
  ground advantage: $\lambda=1$ in Table~\ref{windrawloss}], the
probability of win/draw/loss for Rangers would be
$\left(\frac{p_R}{(1+D)(p_R+p_C)};\frac{D}{1+D};\frac{p_C}{(1+D)(p_R+p_C)}\right)$,
estimated at $(0.39;0.27;0.34)$.
%
<<makecontourdata, echo=FALSE, cache=TRUE>>=
n <- howmany
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
@

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce~\citep{plackett1975} likelihood function for
this observation:
%
\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}
%
Note the use of multiplication between the terms, indicating
conditional independence.  Now we use the multiplicative
generalization of Bradley-Terry strengths to introduce an element of
non-independence between the terms.  There are many ways to do this,
but one simple way would be to define equivalence classes of the
competitors with each equivalence class comprising mutually supporting
runners.  The metaphor would be that a runner who has finished the
race is able to support other members of his equivalence class by
cheering his still-running teammates, boosting their performance.  The
\pkg{hyper2} package includes function \fct{cheering} which
implements this functionality.

@@

Specifically, it seems reasonable to assume that if we ask our
respondent to rank his options, it is highly probable that he will put
RB and BB consecutively (because they are essentially
indistinguishable).  Can we quantify the strength of this effect?  To
do this, we define a bespoke function \fct{RB\_BB\_LF} which
returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options.
%
<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
    )
}
@
%
Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Vector \code{ec} places the options into equivalence classes;
equivalence class 3 comprises the two buses and the other classes are
unsupported as they are of size one.  Vector \code{h} encodes the
strength of the nonindependence.  Observe that in almost every case he
ranks RB and BB consecutively.  Function \fct{RB_BB_LF} takes
argument \code{lambda} that quantifies the perceived similarity
between RB and BB in the same way as function \fct{oldfirm} above.
%
<<definefunctiono>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
@ 
%
Above, we define function \fct{o} which returns the support for a
given value of $\lambda$ at the evaluate [over permissible Bradley
  Terry strengths for the options].  We can treat \fct{o} as a
support function for $\lambda$, and then maximize it:
%
<<label=maxlikesim, cache=TRUE>>=
(osup <- optimize(o, c(10, 40), maximum=TRUE))
@ 
%
So a likelihood ratio test of the null that $\lambda=1$ would be:
%
<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- osup$objective - o(lambda = 1))
@
%
We can thus improve the hypothesis that $\lambda=1$, corresponding to
an unaltered Plackett-Luce system, by over 5.6 units of support by
dint of choosing instead $\lambda = 10.49$.  This easily satisfies
\citeauthor{edwards1992}'s two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
%
%It is straightforward to find a profile
%support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
%We see that support of $\log\lambda$ is closely quadratic over a broad
%range.
%
%\begin{figure}[htbp]
%\begin{center}
<<plotproflike, echo=FALSE>>=
if(FALSE){
    lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
    L <- sapply(lambda, o)
    plot(log(lambda), L-max(L), type="b")
    abline(h=c(0, -2))
    abline(v=0)
}
@ 
%\caption{A\label{profilesupportlambdaRBBB} profile support function
%  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
%  Vertical line indicates null of $\lambda=1$, horizontal lines show
%  maximum support and $-2$}
%\end{center}
%\end{figure}
